---
title: "RMarkdown: Write for Science!"
author: "Michael A Barrowman"
date: "28/04/2020"
knit: (function(...) rmarkdown::render(...,output_format="all"))
output:
  bookdown::word_document2:
    default
  bookdown::pdf_document2:
    default
  bookdown::html_document2:
    default
bibliography: References.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Introduction
Have you ever tried moving a table in Word? How about inserting an image into Google Docs? Switching between your stats program and report becoming a nightmare? Do you hate citations [@Testcite]?

# What can we do?
This course will help you to transition smoothly into the world of RMarkdown, where you can do your statistical analysis and write your entire report/essay in a single document! RMarkdown takes care of things like formatting and figure/table placement for you without worrying about the nitty gritty, see Table \@ref(tab:One) for details

```{r, echo=F}
df <- tibble(x=runif(20),
       y=runif(20))
knitr::kable(head(df),caption="Here's some data!",digits=3,label="One")
```

# What can we make?
RMarkdown also unifies your output: Word, PDF, Webpage, Presentations and Posters can all come out of a singe document with the touch of a button.


```{r, fig.cap="Plots are easy!",echo=F,label="Two",fig.align="center"}
ggplot(df) + geom_point(aes(x,y))
```


# How?
We combine elements of R (the statistical package) and Markdown (the way you format comments on Reddit) into one system and *knit* them into a document. For an example of what you'll be able to produce after this session, see here:

* [HTML](Example.html) A html version of this file
* [PDF](Example.pdf) A pdf version of this file
* [Word](Example.docx) A Word version of this file


# References


